Automatically Splitting a Two-Stage Lambda Calculus

### 2015-10-16 paper/examples.tex
various edits

This is an instance where stage-splitting is more practical than partial
evaluation---if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup, by not expanding some calls to
\texttt{partition}, would result in duplicating first-stage computations.)

@@

Note that the recursive \texttt{partition} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.

@@

Pipeline programs can be directly expressed as staged functions in
\lang\ with splitting used to generate monostaged code required for
each GPU pipeline stage. Composition of such functions in
\lang\ allows more complex pipelines to be expressed in a modular
programming style that is becoming increasingly desirable as graphics
programs increase in complexity\,\cite{Foley:2011,He:2014}.

A simple example of composition of staged functions (representing GPU
pipeline programs) is given below.  The code defines a pipeline as a
staged function that takes an object definition in the first stage
(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
stage and emits the color of the object at the specified pixel. The
code defines a pipeline combinator \texttt{shade} which modulates the
results of the object's reflectance at a given pixel with the results
of a function that computes the object's albedo (surface color).
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Whereas prior systems provides combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

>>>>

This is an instance where stage-splitting is more practical than partial
evaluation---if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup by not expanding some calls to
\texttt{part}, would result in duplicating first-stage computations.)

@@

Note that the recursive \texttt{part} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{part}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.

@@

Pipeline programs can be directly expressed as staged functions in
\lang\ with splitting used to generate monostaged code required for
each graphics pipeline stage. Composition of such functions in
\lang\ allows more complex pipelines to be expressed in a modular
programming style that is becoming increasingly desirable as graphics
programs increase in complexity\,\cite{Foley:2011,He:2014}.

A simple example of composition of staged functions (representing graphics
pipeline programs) is given below.  The code defines a pipeline as a
staged function that takes an object definition in the first stage
(\texttt{obj}) and a pixel coordinate (\texttt{xy}) in the second
stage and emits the color of the object at the specified pixel. The
code defines a pipeline combinator \texttt{shade} which modulates the
results of the object's reflectance at a given pixel with the results
of a function that computes the object's albedo (surface color).
%
\begin{lstlisting} 
1`datatype object = ...`
2`atsignnext{
  type coord = int * int
  type color = int
}`
1`type pipeline = object * $`2`coord`1` -> `$2`color`

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Whereas prior systems provide combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

### 2015-10-16 paper/implementation.tex
various small edits
various edits

@@

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngr{`1`
  datatype list = Empty | Cons of int * list
  fun partition (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
val 3`gr{`partition3`}` = 3`gr{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

>>>>

\subsection {Proof Implementation}

We have encoded the type system, dynamic semantics, splitting algorithm, and output semantics in Twelf,
as well as all of the theorems of \ref{sec:semantics,sec:splitting}, and their proofs. 
Details can be found in the supplementary materials.

\subsection {Example Implementation}

@@

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngr{`1`
  datatype list = Empty | Cons of int * list
  fun part (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
val 3`gr{`part3`}` = 3`gr{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

### 2015-10-16 paper/intro.tex
various edits

In this paper, we present a splitting algorithm for \lang, 
a two-staged typed lambda calculus in the style of Davies\,\cite{davies96},
with support for recursion and first-class functions.
Like Davies, \lang\ uses a $\fut$ modality to denote computation in the second stage,
but to aid splitting we also add a $\curr$ modality to denote purely first-stage computations. 
The dynamic semantics (\ref{sec:semantics}) of \lang\ are that of Davies, 
modified to provide an eager behavior between stages, which is more intuitive in the context of splitting.
We then prove the correctness of our splitting algorithm (\ref{sec:splitting}) for \lang\ with respect to the semantics.
Finally, we discuss our implementation of this splitting algorithm
(\ref{sec:implementation}) and demonstrate its power and behavior
for a number of staged programs ranging from straight-line arithmetic
operations to recursive and higher-order functions (\ref{sec:examples}).

>>>>

In this paper, we present a splitting algorithm for \lang, 
a two-staged typed lambda calculus in the style of Davies\,\cite{davies96},
with support for recursion and first-class functions.
Like Davies, \lang\ uses a $\fut$ modality to denote computation in the second stage,
but to aid splitting we also add a $\curr$ modality to denote purely first-stage computations. 
The dynamic semantics (\ref{sec:semantics}) of \lang\ are that of Davies, 
modified to provide an eager behavior between stages, which we believe is more intuitive in the context of splitting.
We then prove the correctness of our splitting algorithm (\ref{sec:splitting}) for \lang\ with respect to the semantics.
Finally, we discuss our implementation of this splitting algorithm
(\ref{sec:implementation}) and demonstrate its power and behavior
for a number of staged programs ranging from straight-line arithmetic
operations to recursive and higher-order functions (\ref{sec:examples}).

### 2015-10-16 paper/overview.tex
various edits

Suppose that we wish to perform a series of order statistics queries
on a list \texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{partition} depend only on \texttt{l}, although different calls to
\texttt{partition} are made by each call to \texttt{qs}.

@@

In \ref{fig:qs-staged} we show a staged version of \texttt{qs}, called
\texttt{qss}, where first-stage code is colored red, and second-stage blue. In
\texttt{qss}, we regard the input list \texttt{l} as arriving in the first stage
(with type $\curr\mathrm{list}$, a list ``now''), the input rank \texttt{k} as
arriving in the second stage (with type $\fut\rmint$, an integer in the
``future''), and the result as being produced in the second stage (with type
$\fut\rmint$).

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code, and $\texttt{gr}$, signaling ground terms.  
%
We also use a function
\lstinline{1`hold : ^int -> $`2`int`},
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
computations and places the information needed by the subsequent function into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{partition} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

>>>>

Suppose that we wish to perform a series of order statistics queries
on a list~\texttt{l}. To this end, we can use the quickselect
algorithm~\cite{Hoare:1961}, which given a list~\texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
\texttt{qs} partitions \texttt{l} using the first element as
a pivot and then recurs on one of the two resulting sides, depending on
the relationship of \texttt{k} to the size \texttt{i} of the first half, in
order to find the desired element.  
If the index is out of range, a default value of \texttt{0} is returned.
Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permuting the input), \texttt{qs}
runs in expected $O(n)$ time.
%
Using \texttt{qs}, we can perform $m$ different order statistics queries with
ranks $\mathtt{k1},\dots,\mathtt{km}$ as follows:
%
\begin{lstlisting}
(qs l k1, qs l k2, ..., qs l km)
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time.

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{part} depend only on \texttt{l}, although different calls to
\texttt{part} are made by each call to \texttt{qs}.

@@

In \ref{fig:qs-staged} we show a staged version of \texttt{qs}, called
\texttt{qss}, where first-stage code is colored red, and second-stage blue. In
\texttt{qss}, we regard the input list~\texttt{l} as arriving in the first stage
(with type $\curr\mathrm{list}$, a list ``now''), the input rank \texttt{k} as
arriving in the second stage (with type $\fut\rmint$, an integer in the
``future''), and the result as being produced in the second stage (with type
$\fut\rmint$).

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code. Additionally, $\texttt{gr}$ (\emph{ground}) annotations 
mark certain first stage components as being purely first-stage, rather than mixed-stage.
%
We also use a function 
\lstinline{1`hold : ^int -> $`2`int`}, 
\begin{comment}$ this comment block is here for syntax highlighting reasons\end{comment}
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

@@

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this function performs these
computations and places the information needed by the subsequent function into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{part} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

### 2015-10-16 paper/relatedwork.tex
various edits

Researchers have explored other staging transformations that, 
like splitting, partition an input two-stage program into two components,
on corresponding to each stage.  
In particular, \emph{binding time separation}\,\cite{Mogensen89a} 
(also called \emph{program bifurcation}\,\cite{DeNiel91})
has been used as a preprocessor step in partial evaluators,
allowing efficient specialization of programs with 
mixed-stage data structures without changes to the specializer itself.
Notably, the grammar-based binding-time specifications used in binding 
time separation are capable of describing
data structures with purely-static, purely-dynamic, and mixed-stage
content, much like the type system of \lang\
(though this correspondence is less clear without our addition of 
\bbonep\ and $\curr$).

>>>>

Researchers have explored other staging transformations that, 
like splitting, partition an input two-stage program into two components,
one corresponding to each stage.  
In particular, \emph{binding time separation}\,\cite{Mogensen89a} 
(also called \emph{program bifurcation}\,\cite{DeNiel91})
has been used as a preprocessor step in partial evaluators,
allowing efficient specialization of programs with 
mixed-stage data structures without changes to the specializer itself.
Notably, the grammar-based binding-time specifications used in binding 
time separation are capable of describing
data structures with purely-static, purely-dynamic, and mixed-stage
content, much like the type system of \lang\
(though this correspondence is less clear without our addition of 
\bbonep\ and $\curr$).

### 2015-10-16 paper/semantics.tex
various edits

\input{figures/dynamics}

The central tenet of a staged language
is that first-stage code should be evaluated entirely before second-stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input top-level program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since $q$ is monostage by definition, this second pass is standard unstaged evaluation
and is not described in further detail in this paper.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

>>>>

\input{figures/dynamics}
\begin{abstrsyn}
The central tenet of a staged language
is that first-stage code should be evaluated entirely before second-stage code.
Accordingly, our dynamics operates in two passes.
The first pass takes an input top-level program $\coltwo e A$
and reduces all of its first-stage (worlds \bbonem\ and \bbonep) subterms in place,
eventually resulting in a residual $\exv q$.
The second pass further reduces this residual.
Since $q$ is monostage by definition, this second pass is standard unstaged evaluation
and is not described in further detail in this paper.
Moreover, for the purposes of these dynamics, we consider a top level program to always be typed at world \bbtwo.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
expensive computation $e$ depending on how many times $x$ appears in $e$. 
Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}\end{abstrsyn}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
We read this as saying somewhere within $e$ there was the subterm $q$ which needs to be hoisted out,
yielding the new term $e'$ which has a new variable $y$ where the $q$ used to be.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

### 2015-10-16 paper/splitting.tex
various edits

Since \lang\ has three classes of multistage term, 
our splitting algorithm has three forms: 
$\splittwo e A p l r$ for \bbtwo-terms, 
$\splitone e A c {l.r}$ for \bbonem-terms, and 
$v \vsplito \mval i q$ for partial values.
In each form the output has two parts, corresponding to the
first-stage ($p$, $c$, and $i$) and second-stage ($l.r$, and $q$) content of the input.
Note that there's no need to provide forms of splitting for ground terms or residuals,
since those classes of term are already monostage by construction.

@@

An illustrative example is the \texttt{partition} function in the quickselect
example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
then (like \texttt{qs}) it would split into two functions
\texttt{partition1} and \texttt{partition2}, the latter of which recursively
computes the (trivial) second-stage component of \texttt{partition}'s result.
Moreover, \texttt{qs2} would call \texttt{partition2}, just as \texttt{qs1}
calls \texttt{partition1}:
\begin{lstlisting}
2`fun qs2 (p : tree, k : int) : int =
... | Branch (i,h,p1,p2) =>
  let val () = partition2 ((),()) in
  case compare k i of ...
\end{lstlisting}

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  
Defining \texttt{partition} in this world is 
tantamount to proving it has no second-stage computations,
allowing splitting to avoid generating the resumer \texttt{partition2}
and calling it from \texttt{qs2}.

>>>>

Since \lang\ has three classes of multistage term, 
our splitting algorithm has three forms: 
$\splittwo e A p l r$ for \bbtwo-terms, 
$\splitone e A c {l.r}$ for \bbonem-terms, and 
\mbox{$v \vsplito \mval i q$} for partial values.
In each form the output has two parts, corresponding to the
first-stage ($p$, $c$, and $i$) and second-stage ($l.r$, and $q$) content of the input.
Note that there's no need to provide forms of splitting for ground terms or residuals,
since those classes of term are already monostage by construction.

@@

An illustrative example is the \texttt{part} function in the quickselect
example of \ref{sec:exampleQS}. If \texttt{part} were defined at \bbonem\
then (like \texttt{qs}) it would split into two functions
\texttt{part1} and \texttt{part2}, the latter of which recursively
computes the (trivial) second-stage component of \texttt{part}'s result.
Moreover, \texttt{qs2} would call \texttt{part2}, just as \texttt{qs1}
calls \texttt{part1}:
\begin{lstlisting}
2`fun qs2 (p : tree, k : int) : int =
... | Branch (i,h,p1,p2) =>
  let val () = part2 ((),()) in
  case compare k i of ...
\end{lstlisting}

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  
Defining \texttt{part} in this world is 
tantamount to proving it has no second-stage computations,
allowing splitting to avoid generating the resumer \texttt{part2}
and calling it from \texttt{qs2}.
